NumPy---Array Programming in Python

### 2020-01-25 body.tex
Think about expanding interop section

To manipulate the data structure in an intuitive manner, we provide users with
a simple yet powerful array expression syntax.
It supports operations such as indexing (selecting subarrays) and elementwise
calculations.
A high level syntax allows users to express themselves succinctly, while NumPy
deals with the underlying mechanics of making operations fast.

@@

A powerful array feature that affects the way NumPy indexes and combines
arrays, is {\em broadcasting}.
When performing an elementwise operation (such as addition) on two arrays, one
expects that these arrays should have the same shape.
However, through broadcasting, NumPy allows the dimensions to differ, while
still producing results that appeal to intuition.
A trivial example is the addition of a scalar value to an array, but it also
generalizes to more complex examples such as scaling each column of an array,
or generating a grid of coordinates.
In broadcasting, one or both arrays are virtually duplicated (that is, without
copying any data in memory), so that the shapes of the operands match
(Fig.~\ref{fig:array-concepts}d).
Broadcasting is also applied whenever an array is indexed using arrays of
indices (Fig.~\ref{fig:array-concepts}c).

@@

Many of these functions, such as summation, support \emph{reductions}: aggregating
results across one or more dimensions of the array.
For example, summing a $n$-dimensional array over $d$ axes results in a
$(n-d)$-dimensional array (Fig.~\ref{fig:array-concepts}e).

@@

\section*{Supporting the ecosystem}

\fixme{Maybe some wording around the actual *support* element?}

@@

Users predominantly interact with NumPy arrays using {\em indexing} (to access
subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$,
and, for matrix multiplication, $@$), as well as {\em array-aware functions};
together, these provide an easily readable, expressive, high-level API for
array programming.

Various libraries build on and extend this functionality to further enrich the
ecosystem of available tools.  For example, SciPy adds various mathematical,
scientific, and engineering routines; Pandas provides indexed arrays, known as
dataframes; xarray implements arrays with labeled axes; and Matplotlib
generates publication-ready figures and visualizations.  Employing these
low-level array-aware functions are widely used technique-specific libraries,
including scikit-learn (machine learning), scikit-image (image processing),
statsmodels (statistics),  networkx (graphs and complex networks)
\cite{SciPyProceedings_11}, and Napari (volumetric image visualization).

Finally, domain-specific libraries and applications are built on this diverse
foundation of tooling; these include Astropy (astronomy) \cite{astropy:2013,
astropy:2018}, Biopython (computational biology and bioinformatics) \cite{cock2009biopython},
NIPY (neuroimaging) \cite{millman2007analysis}, and Pangeo (earth science) \cite{2018EGUGA..2012146H}.

Exposing array programming primitives, as well as the surrounding ecosystem of
tools, in Python---an interpreted language---creates an ideal environment for
interactive, exploratory data analysis where users may iteratively inspect,
manipulate, and visualize their data \cite{perez2007ipython}.

\section*{The proliferation of arrays}

\fixme{Add references somewhere to Zarr, https://arrow.apache.org/,
https://parquet.apache.org/,
https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}

@@

@@

There has been a steady increase in the number of external libraries that
provide arrays and a NumPy-like API for manipulating them.
This includes popular tensor computation libraries such as
CuPy\footnote{\url{https://cupy.chainer.org/}},
JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}, and
Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
and
Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
provide tensor APIs with NumPy-inspired semantics.
Ideally, operating on these arrays using NumPy would simply work, so that end
users could write code once, and would then benefit from switching between
NumPy arrays, GPU arrays, distributed arrays, and so forth, as appropriate.

@@

To facilitate \emph{interoperability}, the NumPy team has designed
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".

@@

Subclasses of NumPy often exist due to difficulty in constructing richer data
types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint},
geometrical objects \cite{pygeos}, and missing numbers.
To improve \emph{extensibility}, we are currently overhauling the data type
system to make it easier to implement data types both in Python and C, and to
support these and other applications.

@@

% aka why is this so successful
% some sense of ongoing work / future directions

NumPy was initially developed by students, faculty, and researchers in their
spare time to provide a modern array programming library for Python.  They were
influenced by their experiences with powerful interactive programming languages
for scientific computing like Yorick \cite{munro1995using} as well as
commercial languages and environments like IDL and Matlab.  They also
frequently developed code from scratch to solve their own or their colleagues'
problems, often in low-level languages that preceded Python like Fortran
\cite{dongarra2008netlib} and C.

% Fortran / C
% A few sentences on the history of array programming, referencing APL and Fortran, are missing.

This description of how APL users work easily applies to scientific Python
users \cite{kromberg2007apl}:
\begin{quote}
Users of APL tend to ``live in their data''. Interactive interpreters allow them
to inspect, almost feel their  way forward, discovering successful snippets
of code through experimentation and collecting them
into imperative or functional programs according to taste.
\end{quote}

NumPy spurred the development of the larger scientific Python ecosystem and
heralded the current era of wide-spread use of Python for scientific computing.
There were several factors that allowed this rapid growth and successful
development.

Python is already a good choice for standard programming tasks such as
cleaning data, interacting with web resources, and parsing text.
It has a wide range of libraries for different tasks.
Adding fast array operations and linear algebra allows the scientist to do all
their work with within the same language---and one that has the advantage of
being famously easy to learn and teach.

NumPy has strong culture of using software engineering practice to
improve collaboration and reduce error \cite{millman2014developing}.
The NumPy team was early in adopting distributed revision control and code
review to improve collaboration on code, and continuous testing that runs a
large battery of automated tests for every proposed change to NumPy.
The project has comprehensive, high-quality documentation, integrated with the
source code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

@@

This culture of using best practices for producing reliable scientific software
has been eagerly adopted by the ecosystem of libraries that build on NumPy.
For example, in a recent award given by the Royal Astronomical Society to
Astropy, they state:
\begin{quotation}
\noindent\emph{The Astropy Project has provided hundreds of junior scientists
with experience in professional-standard software development practices
including use of version control, unit testing, code review and issue tracking
procedures. This is a vital skillset for modern researchers that is often
missing from formal university education in physics or astronomy.}
\end{quotation}
Community members explicitly work to address this lack of formal education
through formal courses and workshops
\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.
%\fixme{cite https://scipy-school.org/}.

Lastly, NumPy benefited greatly from the simplicity of its underlying data
structure, which made it a natural standard for exchanging array data between
libraries.  It also made it easy for other libraries to develop fast and
memory-efficient compiled code, usually in C or Fortran, that could manipulate
these arrays and pass them back to Python.

Over time, and with the arrival of different technologies and array types, the
role of NumPy has changed from merely being an array provider, to being a
widely adopted array programming language.

% S:
%
% Thinking about enabling factors, I'd characterize them into three categories:
%
% - Practical
% - Philosophical
% - Social
% - Technical
%
% E.g., practical: reason why we use Python (learn one language for
% everything); students don't have money, so want to avoid impracticalities of
% license dongles/servers, etc.
%
% Philosophical: science should be open, transparent; our software should be
% controlled by scientists, not designers that we don't have access to
%
% Social: joy of building these things together, friendly welcome into the
% community for many of us---appreciation of our work and hours

>>>>

Users interact with NumPy arrays using {\em indexing} (to access
subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$
for elementwise operations and $@$ for matrix multiplication), as well as {\em array-aware functions};
together, these provide an easily readable, expressive, high-level API for
array programming, while NumPy
deals with the underlying mechanics of making operations fast.

@@

@@

When performing an elementwise operation (such as addition) on two arrays, one
expects that these arrays should have the same shape.
However, through \emph{broadcasting}, NumPy allows the dimensions to differ, while
still producing results that appeal to intuition.
A trivial example is the addition of a scalar value to an array, but it also
generalizes to more complex examples such as scaling each column of an array,
or generating a grid of coordinates.
In broadcasting, one or both arrays are virtually duplicated (that is, without
copying any data in memory), so that the shapes of the operands match
(Fig.~\ref{fig:array-concepts}d).
Broadcasting is also applied whenever an array is indexed using arrays of
indices (Fig.~\ref{fig:array-concepts}c).

Many of the array-aware functions, such as summation, support \emph{reductions}: aggregating
results across one or more dimensions of the array.
For example, summing a $n$-dimensional array over $d$ axes results in a
$(n-d)$-dimensional array (Fig.~\ref{fig:array-concepts}e).

@@

\section*{Scientific Python ecosystem}

@@

The brevity and expressive power of NumPy's array programming syntax combined with 
an advanced interactive environment like IPython \cite{perez2007ipython}
or Jupyter \cite{Kluyver:2016aa} makes interactive data exploration fluid and simple.
It is common for NumPy users to work interactively with their data,
discovering through experimentation bits of code that they stitch together into
imperative or functional programs.
The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
this solid foundation with core (e.g., SciPy and Matplotlib) libraries,
widely used technique-specific
libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
and several domain-specific projects and libraries
\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
Exposing these simple array programming primitives, as well as the surrounding ecosystem of
tools, in Python---an interpreted, general purpose language---creates an ideal environment for
interactive, exploratory data analysis where users may iteratively inspect,
manipulate, and visualize their data.
This rich environment has made the scientific Python ecosystem popular for scientific
research.

\begin{figure}
  \centering
  \includegraphics[width=.48\textwidth]{static/ecosystem}
  \caption{\textbf{Scientific Python Ecosystem.}
   This figure highlights some of the essential tools and libraries in the vast
   scientific Python ecosystem.  Most of the ecosystem builds on NumPy to
   provide a rich and powerful collection of array-aware functionality.
  }
  \label{fig:ecosystem}
\end{figure}

For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
collaboration relies on many components of the scientific Python ecosystem.
NumPy arrays are used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image 
reconstruction.
Scipy is used to provide tools used in general image processing such as 
filtering and image alignment, while scikit-image provides higher-level
functionality such as edge filters and Hough transforms.
SciPy provides fundamental algorithms for scientific computing on arrays
including many mathematical, scientific, and engineering routines.
SciPy and NumPy are closely related and were developed together often
by the same people.
scikit-image is a popular technique-specific library,
which vastly extends the basic image processing tools provided by SciPy.
Other components of Scipy are used throughout the library, such as the
\code{scipy.optimize} package for handling general optimization tasks.
NetworkX \cite{SciPyProceedings_11} (the standard package for complex
network analysis) is used for analysis of consistency among image
comparisons.
Astropy \cite{astropy:2013, astropy:2018} (a community-developed astronomy
library) is used throughout the library,
primarily for I/O from standard astronomical file formats and time/coordinate
transformations.
Matplotlib (the core library for generating publication-ready figures
and visualizations) is used for visualizing data throughout the analysis pipeline,
including the generation of the final image of the black hole.
% N.B. - most of this comes from introspecting the Image class defined in
% ehtim/image.py and grepping the dependencies in requirements.txt

Tools for analyzing data from gravitational wave observatories such as LIGO
and Virgo are provided by the \code{pycbc} package, which was used in the first
detection of gravitational waves \cite{abbott2016observation} and in 
on-going analysis of data from LIGO and Virgo.
\code{pycbc} make extensive use of the scientific Python ecosystem.
The time-series data from the interferometers are stored in NumPy arrays,
while the \code{scipy.signal} is used to construct and apply filters to the
data.
Matplotlib is used to visualize data at all points along the analysis chain
including raw data from the instrumentation and the time-frequency 
visualization of the ``chirps'' that result from mergers of compact binaries
such as black holes\cite{abbott2016observation} and
neutron stars\cite{abbott2017multi}.

The rich ecosystem of tools building on NumPy and the larger community
of scientists and programmers using those tools mean that NumPy has a
responsibility to maintain considerable stability.
This need creates interesting challenges for NumPy development as the
computational environment scientists are analyzing their data on continues
to evolve and change.

\section*{Proliferation of arrays}

%\fixme{Add references somewhere to Zarr, https://arrow.apache.org/,
%https://parquet.apache.org/,
%https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}

% pandas, xarray, ...

@@

This includes popular tensor computation libraries such as
CuPy, JAX, and Apache MXNet.
PyTorch and Tensorflow provide tensor APIs with NumPy-inspired semantics.
% \footnote{\url{https://cupy.chainer.org/}}
% \footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}
% \footnote{\url{https://numpy.mxnet.io/}}
% \footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
% \footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}

@@

There has been a steady increase in the number of external libraries that
provide arrays and a NumPy-like API for manipulating them.
Ideally, operating on these arrays using NumPy would simply work, so that end
users could write code once, and would then benefit from switching between
NumPy arrays, GPU arrays, distributed arrays, and so forth, as appropriate.

@@

To facilitate \emph{interoperability}, the NumPy team has designed
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".
% Below an attempt to start on this:
Thanks to these developments users of Dask and NumPy arrays are now capable
to scale their computation to the GPU using Dasks CuPy integration with
only minimal code changes.
However, general adoption has not yet been achieved mainly for compatibility
concerns.
Thus, we are actively exploring to address these by revising and extending
these protocols with the aim of achieving a situation where this
interoperability spans the whole scientific Python ecosystem up to and including domain libraries.

@@

Subclasses of NumPy often exist due to difficulty in constructing richer data
types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint},
geometrical objects \cite{pygeos}, and missing numbers. % pandas: categoricals, IP addresses?
To improve \emph{extensibility}, we are currently overhauling the data type
system to make it easier to implement data types both in Python and C, and to
support these and other applications.
The widespread adoption of rich, user defined, data types can allow more
concise, feature rich scientific computing.
By making these user implemented data types a fundamental design of NumPy
these tools will enabled and interact naturally with different libraries.

@@

NumPy was initially developed by students, faculty, and researchers to
provide an advanced, open source array programming library for Python,
which was free to use and unencumbered by license servers, dongles, and the like.
There was a sense of building something
impactful together, for the benefit of many others.  Participating in
such an endeavor, within a welcoming community of like-minded
indivuals, held a powerful attraction for many early contributors.

These user-developers frequently had to write code from scratch to solve
their or their colleagues' problems---often in low-level languages
that preceded Python, like Fortran \cite{dongarra2008netlib} and C.
To them, the advantages of an interactive, high-level array library
were evident. The design of this new tool was informed by their
experiences with powerful interactive programming languages for
scientific computing such as APL \cite{iverson1962programming} and
Yorick \cite{munro1995using}, as well as commercial languages and
environments like IDL and Matlab.

NumPy has subsequently seen prolific adoption, with millions of users,
first from research and later also industry.
It spurred the development of the larger scientific Python ecosystem and
heralded the current era of wide-spread use of Python for scientific computing.
The reasons for its success are varied and complex, but include the fact that
NumPy (a) is built on top of Python,
(b) has a strong culture of collaboration and good software development practices, and
(c) provides a cutting-edge array programming environment for scientific data analysis.

\emph{Python} is an open source, general-purpose, interpreted programming language
well-suited to standard programming tasks such as cleaning data,
interacting with web resources, and parsing text.
Adding fast array operations and linear algebra allows scientist to do all
their work within a single language---and one that has the advantage of
being famously easy to learn and teach.
In fact, Python's adoption as a primary learning language in universities
(e.g., many lower division UC Berkeley courses use Python including
Foundations of Data Science,
The Beauty and Joy of Computing,
Introduction to Computational Thinking with Data,
and The Structure and Interpretation of Computer Programs) has
further popularized NumPy as a tool of choice for modern data science.

Even though NumPy is not part of Python's standard library,
we benefit from a good relationship with the Python developers.
Over the years, the Python language has added new features and
special syntax so that NumPy would have more succinct and 
easy to read array notation.
Since we are not part of the standard library, we are able to
dictate our own release policies and development patterns.
%\fixme{we've gotten a number of features: slicing syntax, implicit tuples,
%complex numbers, buffer protocol, matrix multiplication operator}
%\fixme{but we benefit from separate governance}

NumPy is a community of practice with a strong \emph{culture} of
employing software engineering practice to improve collaboration and
reduce error \cite{millman2014developing}.  This culture is not only
adopted by leaders in the project, but also enthusiastically taught to
newcomers. The NumPy team was early in adopting distributed revision
control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for
every proposed change to NumPy.  The project has comprehensive,
high-quality documentation, integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

@@

This culture of using best practices for producing reliable scientific software
has been eagerly adopted by the ecosystem of libraries that build on NumPy.
For example, in a recent award given by the Royal Astronomical Society to
Astropy, they state:
\begin{quotation}
\noindent\emph{The Astropy Project has provided hundreds of junior scientists
with experience in professional-standard software development practices
including use of version control, unit testing, code review and issue tracking
procedures. This is a vital skill set for modern researchers that is often
missing from formal university education in physics or astronomy.}
\end{quotation}
Community members explicitly work to address this lack of formal education
through formal courses and workshops
\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.

NumPy combines the expressive power of \emph{array programming}, 
the performance of C, and the Python programming language in a mature,
well-tested, well-documented, and community developed library.
It builds on decades of work refining the array programming
syntax from APL through Yorick and others.
It also made it easy for other libraries to develop fast and
memory-efficient compiled code, usually in C or Fortran, that could manipulate
these arrays and pass them back to Python.
Popular array-aware libraries increased the user community and that
in turn led to more libraries adding new functionality
to the scientific Python ecosystem.

Over time the role of the NumPy has evolved. It is no longer just the
foundational array library underlying the scientific Python ecosystem,
but it is also a standard API for tensor computation and a
central coordinating mechanism between array types and technologies.
